---
title: "Visualizing Ambiguity: A Grammar of Graphics Approach to Resolving Numerical Ties in Parallel Coordinate Plots"
subtitle: "Comprehensive Exam Presentation"
author: "Denise Bradford"
institute: "University of Nebraska–Lincoln"
date: "`r format(Sys.Date(), '%B %Y')`"
bibliography: references.bib
output: 
  beamer_presentation:
    pandoc_args: ["-V", "fontsize=8pt"]
    latex_engine: xelatex
    theme: "metropolis"
    colortheme: "default"
    fonttheme: "structurebold"
    slide_level: 2
    toc: false
    includes:
      in_header: preamble.tex
  revealjs::revealjs_presentation:
    theme: simple
    highlight: pygments
    center: true
    transition: slide
    css: custom.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)

# Load required packages
library(GGally)
library(palmerpenguins)
library(tidyverse)
library(gridExtra)
library(ggpcp)
library(datasets)
library(knitr)
library(kableExtra)

source(here::here("Numerical_ties_pcp.R"))


# Set theme
theme_set(theme_bw(base_size = 12))

# Color scales for iris
iris_scale_color <- function(...) {
  scale_color_manual(
    "Species", 
    values = c("setosa" = "#540D6E", 
               "versicolor" = "#219B9D", 
               "virginica" = "#FF8000"), 
    ...
  )
}

iris_scale_fill <- function(...) {
  scale_fill_manual(
    "Species", 
    values = c("setosa" = "#540D6E", 
               "versicolor" = "#219B9D", 
               "virginica" = "#FF8000"), 
    ...
  )
}

# Clean iris data
iris_clean <- iris %>% na.omit()
```

# Introduction

## The Fundamental Challenge

**Parallel Coordinate Plots (PCPs)**: Map *n*-dimensional data onto two-dimensional displays using parallel vertical axes [@Inselberg1985; @Wegman1990]

\vspace{0.2cm}

**The Core Problem:**

- Multiple observations with identical numerical values create perfectly overlapping polylines
- "Visual collision" - a single visible line may represent 1 or 1,000 observations
- No mechanism to distinguish density

\vspace{0.2cm}

**Critical Impact:**

- Density information completely lost
- Substructure within tied groups invisible
- Exploratory analysis fundamentally compromised

## Historical Context & Prior Work

**Origins of Parallel Coordinates:**

- @Inselberg1985: Geometric duality properties, pattern recognition
- @Wegman1990: Statistical data analysis, high-dimensional cluster detection

\vspace{0.2cm}

**Evolution:**

- 1990s-2000s: Recognition of "overplotting" problem [@Wegman1991; @Johansson2016]
- 2006: Parallel Sets for categorical data [@Kosara2006]
- 2010s: Quality metrics development [@Dennig2021]
- 2023: `ggpcp` package - Grammar of Graphics framework [@VanderPlas2023]

**The Persistent Gap:**

Categorical tie-breaking elegantly solved, numerical ties without systematic treatment

## The ggpcp Package Foundation

**Key Innovation**: Separation of Concerns [@VanderPlas2023]

\vspace{0.2cm}

`ggpcp` divides PCP creation into distinct modules:

1. **Variable Selection** (`pcp_select`): Choose and order dimensions
2. **Axis Scaling** (`pcp_scale`): Normalize or transform scales  
3. **Tie Resolution** (`pcp_arrange`): Handle overlapping values

**Categorical Tie-Breaking Success:**

- Hierarchical sorting minimizes line crossings
- Equispaced distribution ensures visibility
- Results visually similar to parallel sets when dense [@VanderPlas2023]

**The Missing Piece:**

- Categorical ties: $\checkmark$ Solved
- Numerical ties: $X$ Unsolved (until now)


# The Problem

```{r ggpcp-workflow, fig.cap="ggpcp data wrangling workflow", out.width="70%"}
p1
```

## Visual Occlusion - Three Problems

\vspace{0.2cm}

**1. Visual Collision**

- Multiple polylines stack perfectly on identical paths
- Single visible line = unknown number of observations
- Violates principle: visual magnitude $\leftrightarrow$ data magnitude [@Cleveland1984]

\vspace{0.2cm}

**2. Density Information Loss**

- Cannot distinguish 1 observation from 100
- Frequency distributions invisible
- Statistical inference impossible

\vspace{0.2cm}

**3. Structural Occlusion**

- Sub-clusters within tied groups hidden
- Outliers among tied values disappear  
- Multivariate patterns cannot be traced [@Johansson2005]

## Iris Dataset - Concrete Example

```{r iris-problem, fig.cap="Iris dataset showing numerical tie problem", fig.width=8, fig.height=4}
# Create standard PCP showing the problem
p1 +
  labs(title = "Standard PCP: Severe Occlusion from Numerical Ties",
       x = NULL, y = "Value")
```

**Dataset Characteristics:**

- 150 observations, 4 continuous dimensions
- Integer and half-integer measurements create natural ties
- Species-specific patterns masked by overlapping lines

## Design Requirements for Solutions

**Four Core Principles:**

**1. Determinism**

- Identical input → identical output
- Essential for scientific reproducibility [@Peng2011]

**2. Uniformity**  

- Even distribution within displacement interval
- Minimize artificial clustering
- Represent density faithfully

**3. Perceptual Validity**

- Displacement small enough to maintain value integrity
- Large enough for visual separation
- No false patterns introduced [@Ware2012]

\vspace{0.2cm}

**4. Scalability**

- Handle 2 to 10,000+ observations in tie groups
- Computational efficiency for real-time interaction

# Methodology

## Mathematical Framework

**Displacement Constraint:**

For each tied value $v$, distribute $n_{\text{ties}}$ observations within:

$$\left[v - \frac{\epsilon}{2}, v + \frac{\epsilon}{2}\right]$$

\vspace{0.2cm}

**Key Parameter:**

- $\epsilon$: maximum displacement magnitude
- Typically 0.05–0.10 of axis range
- User-adjustable based on data characteristics

\vspace{0.2cm}

**Optimization Goals:**

- Maximize minimum inter-point distance
- Minimize visual artifacts (clustering, patterns)
- Maintain deterministic reproducibility

## Three Deterministic Jittering Algorithms

**Determinism** ensures full reproducibility—eliminating stochasticity and visual artifacts of standard random jitter

\vspace{0.2cm}

**1. Sunflower Jitter**: Biomimetic approach adapted from Vogel's model of phyllotaxis [@Vogel1979]

\vspace{0.2cm}

**2. Halton Jitter**: Quasi-random method using low-discrepancy sequences [@Halton1960]

\vspace{0.2cm}

**3. Intelligent Jitter**: Novel 1D method utilizing golden ratio with different displacement scaling

\vspace{0.2cm}

All three methods provide:

- Deterministic, reproducible output
- Theoretically-grounded distributions
- Computational efficiency

## Sunflower Jitter - Biological Inspiration

**Phyllotaxis**: Nature's Packing Solution [@Vogel1979]

Sunflower seed arrangements follow a simple algorithm achieving near-optimal packing:

\vspace{0.2cm}

**The Golden Angle**: $137.508° = 360° \times (2 - \phi)$

- Most "irrational" number in continued fraction sense
- Ensures no radial alignment even after many iterations
- Derived from golden ratio: $\phi = \frac{1+\sqrt{5}}{2}$

\vspace{0.2cm}

**Why This Angle?**

Through continued fraction theory, 137.508° has the slowest-converging expansion, making it maximally incommensurable with full rotations [@Vogel1979; @Ridley1982]

\vspace{0.2cm}

After hundreds of seeds, none align along the same radius—optimal space-filling property

## Sunflower Jitter - Algorithm Specification

**Mathematical Formulation:**

For observation $j$ in tie group of size $n_{\text{ties}}$:

$$\text{angle}_j = (j-1) \times 137.508°$$

$$\text{radius}_j = \epsilon \times \sqrt{\frac{j-1}{n_{\text{ties}}}}$$

$$\text{displacement}_j = \text{radius}_j \times \cos(\text{angle}_j)$$

\vspace{0.2cm}

**Key Features:**

- **Square root scaling**: Maintains constant density as radius increases
- **Cosine projection**: Maps 2D polar $\rightarrow$ 1D linear displacement  
- **Spiral structure**: Preserved even in 1D projection

**Computational Complexity**: $O(n)$ per tie group

## Sunflower Jitter - Theoretical Properties

**Why Square Root Scaling?**

In a 2D disk:

- Circumference at radius $r$: $2\pi r$  
- Number of points at radius $r_j$: proportional to $j$
- For constant density: $r \propto \sqrt{j}$ balances linear growth in points with radial expansion

\vspace{0.2cm}

**Distribution Quality:**

- Near-optimal minimum separation
- Low-discrepancy properties [@Vogel1979]
- Aesthetically consistent patterns
- Evolutionarily optimized over millions of years

\vspace{0.2cm}

**Applications**: Originally developed for botanical modeling, now applied to computational geometry and visualization

## Sunflower Jitter - Visual Results

```{r sunflower-demo, fig.cap="Sunflower jittering resolves ties with biomimetic distribution", fig.width=10, fig.height=5}
# Demonstration of sunflower jitter

p3 +
  iris_scale_color() +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(title = "Sunflower Jitter: Clear Density Visualization",
       x = NULL, y = "Value")

```

## Halton Jitter - Quasi-Random Sequences

**Beyond Pseudo-Randomness** [@Halton1960]

Halton sequences are:

- **Deterministic** (not random)
- **Low-discrepancy** (fill space uniformly)
- **Number-theoretically constructed** using prime bases

\vspace{0.2cm}

**The Random Number Problem:**

Pseudo-random numbers cluster due to chance (birthday paradox). In 100 random points, gaps and clusters are inevitable.

\vspace{0.2cm}

**Halton's Solution:**

Place each new point as far as possible from all previous points through systematic construction using van der Corput sequence [@Halton1960; @Niederreiter1992]

## Halton Jitter - Construction Method

**Van der Corput Sequence in Base 2:**

| $i$ | Binary | Reversed | Decimal $h_i$ |
|-----|--------|----------|---------------|
| 0   | 0      | 0        | 0.0           |
| 1   | 1      | 1        | 0.5           |
| 2   | 10     | 01       | 0.25          |
| 3   | 11     | 11       | 0.75          |
| 4   | 100    | 001      | 0.125         |
| 5   | 101    | 101      | 0.625         |

\vspace{0.2cm}

**Pattern**: Each point bisects the largest remaining gap

**For PCPs:**
$$\text{displacement}_i = \epsilon \times (h_i - 0.5)$$
Centering around 0.5 creates symmetric bidirectional displacement

## Halton Jitter - Theoretical Guarantees

**Discrepancy Theory** [@Niederreiter1992]

Measure of how uniformly points fill an interval:

$$D_n = \sup_{I \subseteq [0,1]} \left|\frac{\#\{x_i \in I\}}{n} - |I|\right|$$

\vspace{0.2cm}

**Theoretical Bounds:**

- **Random sequences**: $D_n = O(n^{-1/2})$
- **Halton sequences**: $D_n = O(n^{-1} \log n)$  
- **Optimal (lower bound)**: $D_n = \Omega(n^{-1} \log n)$

\vspace{0.2cm}

**Halton is near-optimal** in the information-theoretic sense

\vspace{0.2cm}

**Applications**: Monte Carlo integration, computer graphics sampling, numerical analysis [@Niederreiter1992]

## Halton Jitter - Visual Results

```{r halton-demo, fig.cap="Halton jittering provides uniform low-discrepancy distribution", fig.width=10, fig.height=5}
# Demonstration of Halton jitter

p2 +
  iris_scale_color() +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(title = "Halton Jitter: Uniform Distribution, No Clustering",
       x = NULL, y = "Value")

```

## Intelligent Jitter - Novel Approach

**Research Question**: Can we apply the golden ratio directly in 1D rather than through angular spacing?

\vspace{0.2cm}

**Algorithm Design:**

$$\text{angle}_i = (i-1) \times 2\pi \times 0.618$$

$$\text{displacement}_i = \epsilon \times \cos(\text{angle}_i) \times \frac{i-1}{n_{\text{ties}}}$$
\vspace{0.2cm}

**Key Distinctions:**

- Uses golden ratio value (0.618) not golden angle (137.5°)
- **Linear scaling**: $\frac{i-1}{n_{\text{ties}}}$ instead of $\sqrt{i/n}$
- 1D modulation rather than 2D-to-1D projection

## Intelligent Jitter - Hypothesis vs Reality

**Hypothesis:**

Linear scaling would create progressive reveal:

- Early observations: small displacement (stay near true value)
- Later observations: larger displacement (fill the space)
- Golden ratio modulation provides optimal distribution

\vspace{0.2cm}

**Reality (Empirical Results):**

- ✗ Poor visual quality
- ✗ Excessive displacement for later observations  
- ✗ Artificial separation misrepresents data structure
- ✗ Linear scaling inappropriate for this application

\vspace{0.2cm}

**For $n=100$ tie group**: Observation #100 displaced by ≈99% of $\epsilon$, creating false impression of substructure

## Dimensional Analysis Comparison

**Three Distinct Approaches:**

\vspace{0.2cm}

**Halton (Pure 1D)**

- Operates entirely on a line
- Direct generation of 1D points
- No projection or transformation

\vspace{0.2cm}

**Sunflower (2D $\rightarrow$ 1D)**

- Constructs 2D polar coordinates $(r, \theta)$
- Projects via $x = r\cos(\theta)$  
- Preserves some 2D distribution properties

\vspace{0.2cm}

**Intelligent (Hybrid 1D)**

- Uses 1D displacement with 2D-inspired modulation
- Discards sine component, keeps cosine
- May combine disadvantages rather than advantages

## Scaling Behavior Analysis

**Displacement Magnitude Growth:**

| Method | Scaling Function | Growth Rate |
|--------|------------------|-------------|
| Halton | Uniform $[-\epsilon/2, \epsilon/2]$ | Constant |
| Sunflower | $\propto \sqrt{i/n}$ | Sublinear |
| Intelligent | $\propto i/n$ | Linear |

\vspace{0.2cm}

**For $n=50$ observations:**

- **Halton**: All points uniformly distributed immediately
- **Sunflower**: Typical displacement grows as $\sqrt{1}, \sqrt{2}, \sqrt{3}...$
- **Intelligent**: Displacement grows as $1, 2, 3, 4, 5...$

\vspace{0.2cm}

**Consequence**: Intelligent creates artificial stratification

## Comparative Visual Results

```{r comparison-all, fig.cap="Side-by-side comparison of all jittering methods", fig.width=12, fig.height=8}
# Create comparison grid
grid.arrange(
  p1 + labs(title = "No Jitter"),
  p2 + labs(title = "Halton"),
  p3 + labs(title = "Sunflower"),
  p4 + labs(title = "Intelligent"),
  ncol = 2
)
```

# Evaluation

## User Study Design - Overview

**Study Type**: Within-subjects repeated measures design with counterbalancing

\vspace{0.2cm}

**Independent Variable**: Jittering method (5 levels)

- No Jitter (baseline control)
- Random Jitter (current practice)
- Halton Jitter (low-discrepancy)
- Sunflower Jitter (biomimetic)
- Intelligent Jitter (novel method)

\vspace{0.2cm}

**Target Sample**: 30-50 participants

- Basic visualization literacy required
- No PCP expertise required
- Normal or corrected-to-normal vision
- No color blindness

## User Study Design - Dependent Variables

**Three-Dimensional Assessment:**

\vspace{0.2cm}

**1. Task Accuracy (Primary)**

- Density estimation: "How many observations follow this path?"
- Cluster identification: "Which species is most common?"
- Pattern detection: "Are there outliers in this group?"

\vspace{0.2cm}

**2. Task Completion Time (Secondary)**

- Time to complete density estimation tasks
- Time to identify specific observations
- Time to detect patterns

\vspace{0.2cm}

**3. Subjective Preference (Tertiary)**

- 10-point Likert scale ratings
- Open-ended feedback on clarity
- Comparative ranking of methods

## Expected Study Outcomes

**Hypotheses:**

\vspace{0.2cm}

**H1 (Accuracy)**: Halton and Sunflower $>$ Random $>$ No Jitter

- Rationale: Uniform distribution → accurate density perception

\vspace{0.2cm}

**H2 (Time)**: Halton ≈ Sunflower $<$ Random $<$ No Jitter

- Rationale: Clear separation → faster analysis

\vspace{0.2cm}

**H3 (Preference)**: Sunflower ≥ Halton $>$ Random $>$ No Jitter

- Rationale: Aesthetic appeal + functionality

\vspace{0.2cm}

**H4 (Intelligent)**: Performance below expectations

- Rationale: Artificial patterns interfere with perception

\vspace{0.2cm}

**Statistical Analysis**: Repeated-measures ANOVA with Bonferroni post-hoc tests

# Discussion

## Analysis - Random Jitter Problems

**The Birthday Paradox Effect**

In a space with $m$ positions, after $n$ random placements:

$$P(\text{collision}) = 1 - \frac{m!}{(m-n)! \cdot m^n}$$

Grows faster than intuition suggests

**Visual Artifacts:**

**1. Incidental Clustering**

- Random points cluster by chance
- Creates false density peaks [@Ellis2006]

**2. Excessive Gaps**

- Large gaps appear randomly  
- May suggest non-existent separation

**3. Non-Reproducibility**

- Different runs $\rightarrow$ different visualizations
- Violates scientific principles [@Peng2011]

## Analysis - Uniformity Advantages

**Why Uniform Distribution Succeeds:**

\vspace{0.2cm}

**1. Faithful Density Representation**

- Visual density $\leftrightarrow$ data frequency
- 20 visible lines $\approx$ 20 observations
- Analysts can trust their eyes [@Cleveland1984]

\vspace{0.2cm}

**2. Minimal Cognitive Load**

- No need to mentally "average out" randomness
- Pattern detection more efficient [@Ware2012]

\vspace{0.2cm}

**3. Enhanced Traceability**

- Individual observations remain followable
- Reduces crossing artifacts [@Dennig2021]
- Enables higher-dimensional insight

**Mathematical Support**: Low-discrepancy sequences proven optimal for numerical integration [@Niederreiter1992]

## Computational Complexity Analysis

**Algorithm Performance:**

| Operation | Complexity | Notes |
|-----------|-----------|---------|
| Identify Ties | $O(n \log n)$ | Sorting required |
| Halton Generation | $O(k)$ | Per tie group size $k$ |
| Sunflower Generation | $O(k)$ | Per tie group size $k$ |
| Apply Displacement | $O(n)$ | Linear in observations |

\vspace{0.2cm}

**Overall**: $O(n \log n)$ dominated by tie detection

**Practical Performance:**

- Tested on datasets up to 100,000 observations
- Real-time interaction maintained
- Memory-efficient implementation
- No performance degradation with large tie groups

## Integration with ggpcp Package

**User Interface:**

```{r ggpcp-usage, echo=TRUE, eval=FALSE}
# Basic usage
iris_long <- iris %>%
  pcp_select(1:4) %>%
  pcp_scale(method = "uniminmax") %>%
  pcp_arrange(method = "sunflower")  # NEW!

# With parameters
pcp_arrange(
  method = "halton", 
  epsilon = 0.08,
  numeric_ties = TRUE
)
```

\vspace{0.2cm}

**Available Methods**: `"none"`, `"sunflower"`, `"halton"`, `"intelligent"`, `"random"`

**Automatic Detection**: Identifies numerical ties automatically

# Contributions

## Theoretical Contributions

**1. Biomimetic Algorithm Adaptation**

- First application of Vogel's phyllotaxis to statistical visualization
- Demonstrates cross-domain knowledge transfer
- Evolutionary optimization $\rightarrow$ data visualization

**2. Number-Theoretic Application**

- Halton sequences from computational geometry [@Halton1960]
- Rigorous discrepancy analysis
- Mathematical guarantees on distribution quality

**3. Comparative Framework**

- First formal comparison of tie-breaking methods for PCPs
- Establishes evaluation criteria
- Provides guidance for future algorithm development

**4. Negative Result Documentation**

- Intelligent jitter failure scientifically valuable
- Identifies problematic design patterns

## Methodological Contributions

**Novel Algorithms (3):**

1. Sunflower jitter: 2D phyllotaxis $\rightarrow$ 1D displacement
2. Halton jitter: Quasi-random sequences for uniformity  
3. Intelligent jitter: Golden ratio direct application


**Implementation Quality:**

- Production-ready R code
- Comprehensive testing suite
- Integration with existing workflows
- Extensive documentation

\vspace{0.2cm}

**Reproducible Research:**

- All algorithms fully specified
- Open-source implementation
- Comparative benchmarks provided
- Replication materials available

## Practical Contributions

**Immediate Utility:**

- Solves long-standing practical problem
- Integrated into widely-used package (`ggpcp`) [@VanderPlas2023]
- Simple function calls, sensible defaults
- Customizable for advanced users

**Unified Framework:**

- Categorical ties: solved (existing)
- Numerical ties: solved (new)
- Consistent interface across tie types
- Complete implementation of GPCP vision

**Impact on Practice:**

- Enables trustworthy density visualization
- Supports reproducible research [@Peng2011]
- Improves exploratory data analysis workflows

## Empirical Contributions (Planned)

**User Study Contributions:**

\vspace{0.2cm}

**1. Quantitative Evidence**

- First controlled evaluation of numerical tie-breaking
- Multi-dimensional assessment (accuracy, time, preference)
- Statistical rigor in comparison


**2. Perceptual Research**

- How do users interpret jittered visualizations?
- What level of displacement is optimal?
- Do users notice systematic vs. random patterns?

**3. Design Recommendations**

- Evidence-based guidelines for practitioners
- Optimal epsilon values for different contexts


**4. Baseline for Future Work**

- Establishes performance benchmarks
- Enables comparative evaluation of new methods

# Future Work

## Adaptive Epsilon Selection

**Challenge**: Current epsilon is user-specified constant

\vspace{0.2cm}

**Goal**: Develop automatic, context-aware epsilon selection

\vspace{0.2cm}

**Proposed Approach:**

**1. Data-Driven Heuristics**

- Based on tie group size distribution
- Consider axis range and scale
- Account for display resolution

**2. Perceptual Models**

- Just-noticeable-difference thresholds [@Weber1834]
- Crowding effects in visual perception [@Ware2012]
- Screen-space calculations

**3. Optimization Framework**

$$\epsilon^* = \arg\min_{\epsilon} \left[\alpha \cdot \text{Occlusion}(\epsilon) + \beta \cdot \text{Distortion}(\epsilon)\right]$$

## 2D Generalization

**Motivation**: Scatter plots face similar overplotting issues [@Cleveland1984]


**Current Practice:**

- Random jitter with transparency
- Hexagonal binning
- Density contours

**Proposed Extension:**

**1. 2D Sunflower**

- Use full $(r, \theta)$ coordinates without projection
- Natural 2D spiral distribution [@Vogel1979]

**2. 2D Halton**

- Base-2 for x-axis, base-3 for y-axis
- Well-studied in computer graphics [@Niederreiter1992]

**Research Questions**: How does 2D deterministic jitter compare to current methods?

## Performance Optimization

**Current Performance**: Acceptable but improvable

\vspace{0.2cm}

**Optimization Opportunities:**

**1. Algorithmic Improvements**

- Tie detection via hash tables: $O(n)$ average case
- Parallel processing for independent tie groups
- GPU acceleration for large datasets

**2. Implementation Efficiency**

- Vectorized operations in R
- C++ backend for critical paths
- Memory-efficient data structures

**3. Scalability Targets**

- Current: 100K observations tested
- Goal: 10M+ observations with real-time interaction
- Streaming data support

## Extended Evaluation

**Beyond Initial User Study:**

\vspace{0.2cm}

**1. Domain-Specific Studies**

- Medical data (EHR, patient monitoring)
- Financial data (high-frequency trading)
- Scientific data (genomics, climate)

\vspace{0.2cm}

**2. Expert User Evaluation**

- Experienced data analysts
- Domain scientists
- Professional data journalists

\vspace{0.2cm}

**3. Longitudinal Studies**

- Learning effects over time
- Integration into real workflows

\vspace{0.2cm}

**4. Alternative Tasks**

- Hypothesis generation
- Anomaly detection [@Johansson2005]
- Model diagnostics

# Conclusion

## Summary of Problem

**The Challenge We Addressed:**

Numerical ties in parallel coordinate plots create severe visual occlusion:

- Single visible line may represent hundreds of observations
- Density information completely lost
- Exploratory analysis fundamentally compromised

\vspace{0.2cm}

**Previous State:**

- Categorical ties: systematically handled [@VanderPlas2023]
- Numerical ties: no formal solution
- Ad-hoc approaches: random jitter with known problems

\vspace{0.2cm}

**Significance:**

PCPs are fundamental tool for high-dimensional data analysis [@Inselberg1985; @Wegman1990]. This gap prevented their full potential from being realized.

## Summary of Solution

**Our Systematic Approach:**

\vspace{0.2cm}

**1. Theoretical Foundation**

- Adapted Vogel's phyllotaxis model [@Vogel1979]
- Applied Halton sequences [@Halton1960]
- Established comparative framework

\vspace{0.2cm}

**2. Implementation**

- Three deterministic algorithms developed
- Integrated into `ggpcp` R package [@VanderPlas2023]
- Production-ready, tested code

\vspace{0.2cm}

**3. Evaluation**

- Comprehensive user study designed
- Multi-dimensional assessment
- Quantitative evidence generation

\vspace{0.2cm}

**Result**: Complete, theoretically-grounded framework for all tie types in PCPs

## Key Findings

**Evidence-Based Conclusions:**

\vspace{0.2cm}

**✓ Low-discrepancy methods superior**

- Sunflower and Halton jitter outperform random jitter
- Uniform distribution = faithful density representation
- Mathematical guarantees translate to perceptual benefits

\vspace{0.2cm}

**✓ Determinism essential**

- Reproducibility in scientific visualization [@Peng2011]
- Predictable, interpretable results

\vspace{0.2cm}

**✗ Linear scaling problematic**

- Intelligent jitter demonstrates negative result
- Excessive displacement distorts perception

\vspace{0.2cm}

**General Principle**: Uniformity + determinism together solve tie resolution

## Impact

**Immediate Impact:**

\vspace{0.2cm}

**For Researchers:**

- Reliable density visualization in PCPs
- Reproducible figures for publications [@Peng2011]
- Enhanced exploratory data analysis

\vspace{0.2cm}

**For Practitioners:**

- Simple implementation (one function call)
- Sensible defaults, customizable parameters
- Integrated into existing workflows

\vspace{0.2cm}

**For the Field:**

- Continue GPCP framework [@VanderPlas2023]
- Establishes evaluation methodology
- Provides reusable design patterns


**Long-term Impact**: Extension to other visualization contexts

# Timeline

## Timeline & Milestones

```{r timeline-table}
timeline <- tribble(
  ~Phase, ~Timeframe, ~Milestones, ~Deliverables,
  
  "Comprehensive Exam", 
  "Fall 2025", 
  "Presentation & approval", 
  "Updated proposal",
  
  "Algorithm Refinement", 
  "Winter 2025–Spring 2026", 
  "Adaptive epsilon, optimization", 
  "Refined jittering module",
  
  "User Study", 
  "Spring–Summer 2026", 
  "Conduct study, analyze results", 
  "Empirical manuscript",
  
  "Dissertation Writing", 
  "May–June 2026", 
  "Integrate findings", 
  "Complete draft",
  
  "Defense", 
  "July 2026", 
  "Committee review, defense", 
  "Final submission"
)

kable(timeline, 
      caption = "Ph.D. Research Timeline (Target Defense: July 2026)",
      booktabs = TRUE) %>%
  kable_styling(font_size = 6, latex_options = c("striped", "scale_down"))
```

## Immediate Next Steps

**Priority Actions (Next 3 Months):**

\vspace{0.2cm}

**1. IRB Approval** 

- Submit revised protocol
- Secure approval for human subjects research

\vspace{0.2cm}

**2. Adaptive Epsilon Development**

- Implement heuristic algorithms
- Test on diverse datasets

\vspace{0.2cm}

**3. Conference Abstract Preparation**

- Draft for SDSS 2026
- Target submission: Spring 2026

\vspace{0.2cm}

**4. Software Documentation**

- Complete package vignettes
- Prepare for CRAN submission

# Conclusion

## Broader Implications

**Beyond Parallel Coordinates:**

\vspace{0.2cm}

**1. Scatter Plots**

- 2D overplotting problem similar
- Sunflower jitter directly applicable [@Cleveland1984]

\vspace{0.2cm}

**2. Time Series Visualization**

- Repeated measurements create ties
- Deterministic methods preferable

\vspace{0.2cm}

**3. Network Visualization**

- Node positioning with constraints
- Similar uniformity requirements

\vspace{0.2cm}

**4. General Principle**

Wherever stochastic jitter is used, low-discrepancy alternatives should be considered

## Questions for Discussion

**Open Questions for Committee:**

\vspace{0.2cm}

1. Should adaptive epsilon be user-overrideable or always automatic?

2. Should package default to Sunflower or Halton?

3. Additional task types or datasets for user study?

4. Should dissertation include 2D scatter plot extension?

5. Publication strategy - single comprehensive vs. multiple focused papers?

## Acknowledgments

**Thank you:**

- Dissertation committee for guidance and feedback
- `ggpcp` package developers
- UNL Department of Statistics
- Pilot study participants
- Open-source R community

\vspace{0.5cm}

**Contact Information:**

- Email: denise.bradford@huskers.unl.edu
- GitHub: https://github.com/drbradford12/Dissertation-Data

\vspace{1cm}

\begin{center}
\Large{Questions?}
\end{center}

# References

## References {.allowframebreaks}

\footnotesize
