---
title: "Visualizing Ambiguity: A Grammar of Graphics Approach to Resolving Numerical Ties in Parallel Coordinate Plots"
author: "Denise Bradford"
format: 
  pdf:
    classoption: ["11pt", "english", "singlespacing", "headsepline"]
    geometry:
        - paper=letterpaper
        - inner=2.5cm
        - outer=3.8cm
        - bindingoffset=.5cm
        - top=1.5cm
        - bottom=1.5cm
    include-in-header:
      text: |
        \usepackage{soul}
        \usepackage[dvipsnames]{xcolor} % colors
        \newcommand{\db}[1]{{\textcolor{Green}{#1}}}
        \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
        \newcommand{\hh}[1]{{\textcolor{blue}{#1}}}
  html:
    css: edits.css
bibliography: thesis.bib
bibliographystyle: apa
filters:
   - latex-environment
commands: [db,svp,hh]
---


## The Challenge of High-Dimensional Data

Making sense of complex, multidimensional data is a fundamental challenge in science and industry.

::: {.panel-tabset}
#### The Goal

We want to find patterns, identify clusters, and detect outliers.

```{r goal-plot, echo=FALSE, message=FALSE, warning=FALSE}
library(ggpcp)
library(tidyverse)
data(flea, package = "GGally")

flea %>%
  pcp_select(species, c(2:7), species) %>%
  pcp_scale(method="uniminmax") %>%
  pcp_arrange() %>%
  ggplot(aes_pcp()) + 
    geom_pcp_axes() + 
    geom_pcp(aes(colour = species))

```
#### The Reality

Real-world data is often messy, leading to visualizations that hide more than they reveal.

```{r reality-plot, echo=FALSE, message=FALSE, warning=FALSE}
# Placeholder for a messy, overplotted graph
library(GGally)
data(penguins, package = "palmerpenguins")

penguins %>%
  pcp_select(species, c(3, 4, 5, 6), species) %>%
  pcp_scale(method="uniminmax") %>%
  pcp_arrange() %>%
  ggplot(aes_pcp()) + 
    geom_pcp_axes() + 
    geom_pcp(aes(colour = species)) 

```
:::
---

## A Powerful Tool: Parallel Coordinate Plots (PCPs)

PCPs are designed for this challenge, mapping each data point to a line that traverses parallel axes.

```{r pcp-example, echo=FALSE, fig.align='center', out.width='80%', fig.cap="A clean PCP showing clear separation between groups."}
# A good, clean example of a PCP
# ggparcoord(iris, columns = 1:4, groupColumn = "Species")
knitr::include_graphics("[https://i.imgur.com/y1vJb2F.png](https://i.imgur.com/y1vJb2F.png)") # Placeholder image
```

---

## The Problem: When Data Isn't Unique

PCPs break down with **numerical ties**, which are common with:
-   Categorical variables (e.g., 'low', 'medium', 'high')
-   Integer-based or discretized continuous data

The result is overplotting that completely obscures patterns.

```{r pcp-problem, echo=FALSE, fig.align='center', out.width='80%', fig.cap="An unusable PCP where ties create an uninformative mesh of lines."}
# An example PCP showing the tying/overplotting problem

```

---

## State of the Art: A Grammatical Debate

The **Grammar of Graphics** (Wilkinson, 2005; Wickham, 2010) provides a framework for graphics, but how to handle ties within it is an open question.

:::{.columns}
::::{.column width="50%"}
**The Minimalist View**

-   A graphic is a direct map of data.
-   Overlapping lines *are* the "truth" of the data.
-   Any separation is an artificial manipulation.
-   The burden is on the analyst to know their data.
::::

::::{.column width="50%"}
**The Pragmatic View**

-   The goal of a graphic is *insight*.
-   Overlapping lines destroy insight.
-   The grammar should be flexible enough to resolve visual ambiguity.
-   This view is supported by recent work (VanderPlas et al., 2023).
::::
:::

---

## The Research Gap

While solutions exist (e.g., the `ggpcp` package), significant gaps remain:

. . .

1.  **No Formalism:** How should tie-breaking be formally integrated into the Grammar of Graphics? Is it a new `stat`, `position`, or `scale`?

. . .

2.  **Lack of Perceptual Validation:** We don't know how these modified plots affect human perception and analytical accuracy. Do they help, or do they mislead?

. . .

3.  **Limited Algorithmic Exploration:** Are simple spreading or jittering techniques optimal? Can we design better algorithms that preserve data density?

. . .

4.  **No Practical Guidance:** Practitioners have no evidence-based heuristics to choose the best method for their specific dataset.

---

## My Proposed Research

### Primary Research Question

**How can the formal structure of the Grammar of Graphics be extended to systematically incorporate and evaluate methods for resolving numerical ties in parallel coordinate plots, and what is the quantifiable impact of these methods on the accuracy and efficiency of visual data analysis?**

---

## Secondary Question 1: Theory

**How can the management of numerical ties be most effectively and coherently formalized within the layered grammar of graphics?**

* **Methodology:** Theoretical analysis, literature synthesis, and specification of a new or extended grammatical element.

---

## Secondary Question 2: Methodology

**What are the optimal algorithmic criteria for ordering and spacing tied data points to maximize visual clarity while preserving underlying data properties?**

* **Methodology:** Algorithm design and implementation in R. Develop a novel method that balances separation and density preservation.

---

## Secondary Question 3: Perception

**How do different visualization strategies for ties affect an analyst's ability to perform key visual tasks (cluster identification, outlier detection)?**

* **Methodology:** Controlled user studies with human subjects. Measure task accuracy and completion time across different visualization conditions.

---

## Secondary Question 4: Practice

**Can a set of evidence-based heuristics be developed to guide practitioners in selecting the most appropriate visualization method for their specific data context?**

* **Methodology:** Synthesize findings from RQ1-3 into a practical framework. Validate with case studies on real-world datasets from different domains.

---

## Research Plan & Methodology

A multi-stage approach connecting theory to practice.

```{r mermaid-flowchart, echo=FALSE}
# Using Mermaid for a simple flowchart
library(DiagrammeR)
mermaid("
graph TD
    A[RQ1: Formalize in GoG] --> B(RQ2: Design New Algorithm);
    B --> C{RQ3: Perceptual User Study};
    A --> C;
    C --> D[RQ4: Develop Practical Heuristics];
")
```

The core of the validation will be a **within-subjects user study** comparing task performance across:
1.  **Control:** Traditional PCP with overplotting.
2.  **Baseline:** Standard jittering/spreading (`ggpcp`).
3.  **Proposed:** My novel algorithm from RQ2.

---

## Expected Contributions

This dissertation will contribute new knowledge at four levels:

. . .

1.  **Theoretical:** A formal extension to the Grammar of Graphics for handling data ambiguity.

. . .

2.  **Methodological:** A new, open-source algorithm for visualizing ties in high-dimensional data.

. . .

3.  **Empirical:** The first quantitative evidence of how different tie-breaking methods affect user perception and analytical accuracy.

. . .

4.  **Practical:** Evidence-based guidelines that help researchers and analysts create more effective and honest data visualizations.

---

## Proposed Timeline

A 3-year plan to complete the proposed research.

| Year | Sem 1 | Sem 2 |
| :--- | :--- | :--- |
| **1**| RQ1: Lit Review & Theory | RQ2: Algorithm Prototyping |
| **2**| RQ3: User Study Design & IRB | RQ3: Run Study & Analyze Data |
| **3**| RQ4: Case Studies & Write-up | Dissertation Defense |

---

## Summary

-   **Problem:** Numerical ties make Parallel Coordinate Plots ineffective, hiding critical patterns in data.
-   **Gap:** Existing solutions lack formal grammatical integration, perceptual validation, and practical guidance.
-   **Proposal:** A multi-part research plan to formalize, implement, and evaluate a novel approach to visualizing ties within the Grammar of Graphics.
-   **Impact:** More powerful, reliable, and perceptually-validated tools for high-dimensional data analysis.

---

## Thank You

### Questions?

