---
title: "Visualizing Ambiguity: A Grammar of Graphics Approach to Resolving Numerical Ties in Parallel Coordinate Plots"
subtitle: "Comprehensive Exam Presentation"
author: "Denise Bradford"
institute: "University of Nebraska–Lincoln"
date: "`r format(Sys.Date(), '%B %Y')`"
bibliography: references_summary.bib
output: 
  beamer_presentation:
    pandoc_args: ["-V", "fontsize=8pt"]
    latex_engine: xelatex
    theme: "metropolis"
    colortheme: "default"
    fonttheme: "structurebold"
    slide_level: 2
    toc: false
    includes:
      in_header: preamble.tex
  revealjs::revealjs_presentation:
    theme: simple
    highlight: pygments
    center: true
    transition: slide
    css: custom.css
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)

# Load required packages
library(GGally)
library(palmerpenguins)
library(tidyverse)
library(gridExtra)
library(ggpcp)
library(datasets)
library(knitr)
library(kableExtra)
library(ggforce)

source(here::here("Numerical_ties_pcp.R"))


# Set theme
theme_set(theme_bw(base_size = 11))

# Color scales for iris
iris_scale_color <- function(...) {
  scale_color_manual(
    "Species", 
    values = c("setosa" = "#540D6E", 
               "versicolor" = "#219B9D", 
               "virginica" = "#FF8000"), 
    ...
  )
}

iris_scale_fill <- function(...) {
  scale_fill_manual(
    "Species", 
    values = c("setosa" = "#540D6E", 
               "versicolor" = "#219B9D", 
               "virginica" = "#FF8000"), 
    ...
  )
}

# Clean iris data
iris_clean <- iris %>% na.omit()

img_van_der <- here::here("comprehensive_exam_files", "images", "Van_der_corput_distribution.jpeg")
img_daisy <- here::here("comprehensive_exam_files", "images", "daisy.jpeg")
img_vdc <- here::here("comprehensive_exam_files", "images","vdc_sequence.jpeg")
img_golden_angle <- here::here("comprehensive_exam_files", "images","Phyllotaxis_golden_angle.jpeg")
img_golden_ratio <- here::here("comprehensive_exam_files", "images","TheGoldenRationGraphic.jpeg")

```

# Introduction

## The Fundamental Challenge

**Parallel Coordinate Plots (PCPs)**: Map *n*-dimensional data onto two-dimensional displays using parallel vertical axes [@inselberg1985; @wegman1990]

\vspace{0.2cm}

**The Core Problem:**

- Multiple observations with identical numerical values create perfectly overlapping polylines
- "Visual collision" - a single visible line may represent 1 or 1,000 observations
- No mechanism to distinguish density or trace individual observations

\vspace{0.2cm}

**Critical Impact:**

- Density information completely lost
- Substructure within tied groups invisible
- Exploratory analysis fundamentally compromised
- Cluster identification impossible when ties obscure structure

## Research Context & Gap

**Origins of Parallel Coordinates:**

- @inselberg1985: Geometric duality properties
- @wegman1990: Statistical data analysis, high-dimensional visualization

\vspace{0.2cm}

**Evolution:**

- 1990s-2000s: Overplotting recognized as critical problem [@johansson2016]
- 2003: Hammock plots introduce alternative approach [@schonlau2003]
- 2023: `ggpcp` package - Grammar of Graphics framework [@vanderplas2023]

\vspace{0.2cm}

**The Persistent Gap:**

- Categorical ties: Elegantly solved through hierarchical sorting
- **Numerical ties: No systematic, reproducible solution**

## The ggpcp Package Foundation

**Key Innovation**: Separation of Concerns [@vanderplas2023]

\vspace{0.2cm}

`ggpcp` divides PCP creation into distinct modules:

1. **Variable Selection** (`pcp_select`): Choose and order dimensions
2. **Axis Scaling** (`pcp_scale`): Normalize or transform scales  
3. **Tie Resolution** (`pcp_arrange`): Handle overlapping values

\vspace{0.2cm}

**Categorical Tie-Breaking Success:**

- Hierarchical sorting ("from-left", "from-right") minimizes line crossings
- Equispaced distribution ensures visibility
- Enables individual observation tracing

\vspace{0.2cm}

**The Missing Piece:**

- Categorical ties: Solved
- **Numerical ties: Unsolved until now**

# Key Terminology and Definitions

## Numerical Ties

**Definition:**

Numerical ties occur when multiple observations share identical numerical values on one or more dimensions.

\vspace{0.2cm}

**Sources of Ties:**

- Measurement precision limitations (rounding)
- Discrete measurement instruments
- Natural clustering at specific values
- Data collection procedures

\vspace{0.2cm}

**Distinction from Categorical Ties:**

- Categorical: Expected within groups
- Numerical: May arise from various data characteristics
- Both require treatment but through different mechanisms

## Jittering and Determinism

**Jittering:**

Controlled displacement technique that adds small offsets to prevent visual overlap

- Magnitude controlled by parameter $\epsilon$ (epsilon)
- Purpose: Visual separation, not data modification
- Constraint: Displacement must not mislead about underlying values

\vspace{0.2cm}

**Deterministic vs. Stochastic Methods:**

**Deterministic methods** produce identical results given identical input

- Examples: Halton sequences, Sunflower patterns
- Advantage: Scientific reproducibility
- Requirement: No random number generation

**Stochastic methods** incorporate randomness

- Example: Standard random jitter with `runif()` or `rnorm()`
- Disadvantage: Non-reproducible, may create artifacts

## "Best For" Interpretation

**Definition:**

When we state a method is "best for" a scenario, we mean it optimally balances:

1. **Task Requirements**: What the analyst needs to accomplish
2. **Data Characteristics**: Properties affecting method performance
3. **Performance Criteria**: Quantitative and qualitative measures
4. **Trade-off Optimization**: Aligning method strengths with scenario needs

\vspace{0.2cm}

**Example Applications:**

- **Halton** is best for precision-critical work requiring mathematical guarantees
- **Sunflower** is best for general purpose use balancing uniformity and aesthetics
- **Hammock plots** are best for density visualization over individual tracing

\vspace{0.2cm}

**Important Caveat:**

Recommendations based on theoretical analysis, empirical evaluation, and practical experience. Users should consider their specific context.

# Background and Motivation

## The Problem: Visual Overlap

```{r problem-demo, fig.cap="Standard PCP showing severe numerical tie occlusion"}
p1
```

**Three Critical Issues:**

1. **Visual Collision**: Perfect overlap masks observation count
2. **Density Information Loss**: Cannot distinguish 1 from 100 observations
3. **Structural Occlusion**: Sub-clusters and patterns hidden

\vspace{0.2cm}

**Impact on Analysis:**

- Cluster identification compromised [@blumenschein2020]
- Outlier detection impossible
- Pattern tracing fundamentally limited

## Existing Solutions: Categorical Ties in ggpcp

**ggpcp's Approach:**

Hierarchical sorting through `pcp_arrange(data, method, space)`:

- **"from-left"**: Tie breaks determined by values from left
- **"from-right"**: Tie breaks determined by values from right
- **space parameter**: Controls spacing between categorical levels (default: 5%)

\vspace{0.2cm}

**Key Benefits:**

- Reduces line crossings
- Enables observation tracing
- Provides "external cognition" reducing cognitive load

\vspace{0.2cm}

**Why This Works for Categories:**

- Discrete nature allows hierarchical ordering
- Equispaced distribution natural for categories
- Visual similarity to Parallel Sets when dense

## Alternative Approach: Hammock Plots

**Hammock Plot Strategy** [@schonlau2003; @schonlau2024]:

Uses boxes (parallelograms) where width $\propto$ number of observations

\vspace{0.2cm}

**How Hammock Plots Handle Ties:**

- **Aggregation through width**: Multiple tied observations $\rightarrow$ wider boxes
- **Density through visual magnitude**: Box width directly encodes frequency
- **No separation needed**: Aggregation eliminates occlusion problem

\vspace{0.2cm}

**Advantages:**

- Explicit density representation
- No occlusion with different frequencies
- Seamless handling of mixed variable types

\vspace{0.2cm}

**Limitations:**

- Loss of individual observation tracing
- Increased white space with frugal spacing
- Binning often required for continuous variables

## Comparison: Hammock vs. GPCP

\footnotesize

| Feature | Hammock Plot | GPCP (ggpcp) |
|---------|-------------|--------------|
| Between numerical variables | Constant-width boxes | Lines (overlap) |
| Categorical to numerical | Constant-width boxes | Triangular shapes |
| Individual tracing | Requires highlighting | Natural |
| Density visualization | Explicit (width) | Implicit (overlap) |
| Small datasets | Less detailed | Shows individuals |
| Large datasets | Clearer aggregation | Appears as areas |

\normalsize

\vspace{0.2cm}

**When Hammock Plots Excel:**

- Emphasis on bivariate relationships
- Datasets with many observations per value
- Focus on aggregate patterns over trajectories

## Why ggpcp Needs a Different Solution

**Complementary Approaches:**

Despite hammock plots' success, ggpcp requires numerical tie-breaking:

\vspace{0.2cm}

1. **Preservation of Individual Traceability**: Core ggpcp feature
2. **Grammar of Graphics Philosophy**: Position adjustment fits naturally
3. **Flexibility**: Users choose aggregation or separation based on needs
4. **Small to Medium Datasets**: Where individual observations matter

\vspace{0.2cm}

**Integration Opportunity:**

Future work could combine:

- Jittering for position separation (x-coordinate)
- Width encoding for density (visual weight)
- Benefits: Individual traceability + explicit density

# Design Requirements

## Four Core Principles

**1. Determinism**

Requirement: Identical input must produce identical output

- Rationale: Essential for scientific reproducibility [@peng2011]
- Implication: Rules out standard random jitter

\vspace{0.2cm}

**2. Uniformity**

Requirement: Even distribution within displacement interval with minimal clustering

- Rationale: Faithful density representation, minimize artificial patterns
- Implication: Deterministic methods can guarantee uniform coverage

## Four Core Principles (continued)

**3. Perceptual Validity**

Requirement: Balance two competing needs

- Small enough to maintain value integrity
- Large enough for visual separation
- Rationale: Users must trust displaced values [@ware2012]

\vspace{0.2cm}

**4. Scalability**

Requirement: Handle 2 to 10,000+ observations per tie group efficiently

- Rationale: Real-world datasets vary enormously
- Interactive exploration requires real-time performance
- Implication: Algorithms must have favorable complexity $O(n)$ per group

# Empirical Evidence and Perceptual Foundations

## Perceptual Challenges in Parallel Coordinates

**The Line Width Illusion:**

Users perceive distance between parallel lines at right angles, not vertical distance [@hofmann2013]

- Part of Müller-Lyer family of illusions
- Affects interpretation of separation in PCPs

\vspace{0.2cm}

**Implication for Tie Resolution:**

- Users judge separation based on orthogonal (perpendicular) distance
- Epsilon parameter must account for perceptual bias
- May require different values depending on line angle

\vspace{0.2cm}

**The Sine Illusion:**

Equal-length vertical lines in sine wave pattern appear unequal [@day1991]

- Lines at peaks/troughs appear longer
- Regular patterns can trigger illusion
- Sunflower's golden angle helps prevent periodic patterns

## Clutter and Overplotting: The Core Problem

**Severity:**

Even medium-sized datasets suffer from overplotting, resulting in displays too cluttered to perceive trends or structure [@johansson2016]

\vspace{0.2cm}

**How Ties Exacerbate the Problem:**

- Without ties: Overplotting from similar ranges
- With ties: Perfect overlap of multiple observations
- Result: Complete occlusion with no frequency information

\vspace{0.2cm}

**Existing Clutter Reduction Approaches:**

1. **Clustering-based**: Bands, envelopes, frequency representations
   - Limitation: Loss of individual tracing
2. **Transparency/density**: Alpha blending, density plots
   - Limitation: Fails with perfect overlap (ties)
3. **Our contribution**: Deterministic jittering
   - Resolves ties before rendering
   - Preserves individual traces
   - Complements other methods

## Dimension Ordering Effects

**Critical Importance:**

Order and arrangement of dimensions crucial for PCP effectiveness [@peng2004; @blumenschein2020]

- Similar dimensions should be adjacent
- High impact on visualization quality
- Problem is NP-complete, requires heuristics

\vspace{0.2cm}

**Interaction with Tie Resolution:**

Dimension ordering affects which ties become visible:

```
Ordering 1: A - B - C
  - Ties between A-B highly visible
Ordering 2: A - C - B  
  - Different tie patterns emerge
```

\vspace{0.2cm}

**Implication:**

- Tie detection must be axis-pair specific
- Evaluation must consider multiple orderings
- Integration with existing `pcp_select()` maintains flexibility

## Cluster Identification Performance

**Empirical Findings:**

Recent studies evaluated cluster identification in PCPs [@holten2010; @blumenschein2020]

- Optimal configurations depend on task type and cluster characteristics
- Reordering strategies significantly impact performance

\vspace{0.2cm}

**Relevance to Tie-Breaking:**

Critical question: Does tie-breaking help or hinder cluster identification?

**Potential Positive Effects:**

- Reveals hidden clusters within tied groups
- Improves cluster boundary visibility
- Enables size estimation

**Potential Negative Effects:**

- Displacement might obscure tight clusters
- Cognitive load from visual complexity
- Poor jittering could suggest false clusters

## Task-Dependent Performance

**General Finding:**

PCP effectiveness varies by task complexity [@johansson2016]

\vspace{0.2cm}

**Task Categories for Evaluation:**

\footnotesize

| Task | Difficulty | Expected Impact |
|------|-----------|-----------------|
| **Density Estimation** | Simple | Large improvement with clear separation |
| **Cluster Identification** | Medium | Reveals structure, sensitive to method |
| **Outlier Detection** | Complex | Essential for individual line tracing |

\normalsize

\vspace{0.2cm}

**Integration with Existing Evidence:**

- Use validated tasks from literature
- Control for confounds (ordering, size, design)
- Measure multiple outcomes (accuracy, time, confidence)
- Compare against baselines (No Jitter, Random Jitter)

## Practical Design Recommendations

**Current Best Practices** [@johansson2016; @blumenschein2020]:

1. Manage visual clutter
2. Optimize dimension ordering
3. Consider perceptual factors
4. Support interaction
5. Use appropriate encodings

\vspace{0.2cm}

**Our Extension - Adding a Sixth Principle:**

6. **Resolve numerical ties deterministically:**
   - Apply jittering to prevent perfect overlap
   - Use low-discrepancy methods (Halton/Sunflower)
   - Integrate with existing capabilities
   - Maintain reproducibility

\vspace{0.2cm}

**Implementation in ggpcp:**

All six principles addressed through integrated approach

# Mathematical Framework

## Displacement Constraint

**For each tied value $v$ with $n_{\text{ties}}$ observations:**

Distribute points within displacement interval:

$$\left[v - \frac{\epsilon}{2}, v + \frac{\epsilon}{2}\right]$$

\vspace{0.2cm}

**Key Parameter:**

- $\epsilon$: maximum displacement magnitude
- Typically 0.05–0.10 of axis range
- User-adjustable based on data characteristics and perceptual requirements

\vspace{0.2cm}

**Optimization Goals:**

1. Maximize minimum inter-point distance (prevent collision)
2. Minimize visual artifacts (avoid clustering and false patterns)
3. Maintain deterministic reproducibility (enable verification)
4. Achieve uniform coverage (faithfully represent density)

# Three Deterministic Jittering Methods

## Method Comparison Overview

\footnotesize

| Method | Theoretical Basis | Dimension | Scaling | Best For |
|--------|------------------|-----------|---------|----------|
| **Halton** | Quasi-random sequences [@halton1960] | Pure 1D | Constant | Uniform distributions |
| **Sunflower** | Phyllotaxis [@vogel1979] | 2D $\rightarrow$ 1D | Sublinear $(\sqrt{n})$ | Aesthetic + performance |
| **Intelligent** | Golden ratio direct application | Hybrid 1D | Linear | Research comparison |

\normalsize

\vspace{0.2cm}

**All methods provide:**

- Deterministic, reproducible output
- Theoretically-grounded distributions
- Computational efficiency $O(n)$ or $O(n \log n)$

## Method 1: Sunflower Jitter - Biological Inspiration

**Phyllotaxis**: Nature's optimal packing solution

Sunflower seed arrangements follow evolutionary optimization [@vogel1979]

\vspace{0.2cm}

**The Golden Angle**: $137.508^\circ = 360^\circ \times (2 - \phi)$

where $\phi = \frac{1+\sqrt{5}}{2} \approx 1.618$ is the golden ratio

\vspace{0.2cm}

**Why This Angle Works:**

- Most "irrational" number in continued fraction sense
- Ensures no radial alignment even after hundreds of iterations
- Optimal space-filling property

$$\phi = 1 + \frac{1}{1 + \frac{1}{1 + \frac{1}{1 + \cdots}}}$$

## Sunflower Jitter - Mathematical Formulation

**For observation $j$ in tie group of size $n_{\text{ties}}$:**

$$\text{angle}_j = (j-1) \times 137.50^\circ$$

$$\text{radius}_j = \epsilon \times \sqrt{\frac{j-1}{n_{\text{ties}}}}$$

$$\text{displacement}_j = \text{radius}_j \times \cos(\text{angle}_j)$$

\vspace{0.2cm}

**Key Features:**

- **Square root scaling**: Maintains constant density as radius increases
- **Cosine projection**: Maps 2D polar $\rightarrow$ 1D linear displacement
- **Spiral structure**: Preserved even in 1D projection
- **Biomimetic**: Leverages millions of years of natural selection

## Sunflower Jitter - Why Square Root Scaling?

**Geometric Justification:**

In a 2D disk:

- Circumference at radius $r$: $2\pi r$
- Number of points at radius $r_j$: proportional to $j$
- For constant area density: Need $\frac{dN}{dA} = \text{constant}$

\vspace{0.2cm}

**Mathematical Derivation:**

$$A \propto r^2 \implies N \propto r^2 \implies r \propto \sqrt{N}$$

Therefore: $r \propto \sqrt{j}$ balances linear growth in points with radial expansion

\vspace{0.2cm}

**Distribution Quality:**

- Near-optimal minimum separation [@vogel1979]
- Low-discrepancy properties in 2D
- Aesthetically consistent across scales
- Progressively validated through evolution

## Sunflower Jitter - Applications Beyond PCPs

**Proven Applications:**

\vspace{0.2cm}

- **Point cloud sampling**: Uniform distribution on discs/spheres
- **Sphere packing**: Near-optimal packing density
- **Texture synthesis**: Organic, non-repetitive patterns
- **Computer graphics**: Stratified sampling for ray tracing
- **Quasi-Monte Carlo methods**: Numerical integration

\vspace{0.2cm}

**Why It Works Broadly:**

The golden angle property—maximal incommensurability—creates optimal spacing in any radial system

## Method 2: Halton Jitter - Beyond Pseudo-Randomness

**The Random Number Problem:**

Pseudo-random numbers inevitably cluster (birthday paradox)

\vspace{0.2cm}

**Example:** 100 random points on $[0,1]$:

- Expected maximum gap: 0.05
- Expected minimum gap: 0.0001
- Creates misleading visual artifacts
- Density perception unreliable

\vspace{0.2cm}

**Halton's Solution** [@halton1960]:

Place each new point to systematically fill largest gaps using van der Corput sequence

- Deterministic (not random)
- Low-discrepancy (uniform space-filling)
- Number-theoretically constructed using prime bases
- Mathematically guaranteed coverage

## Van der Corput Sequence Construction

**Algorithm (base 2):**

1. Take integer index $i$
2. Convert to binary
3. Reverse the binary digits
4. Interpret as binary fraction

\vspace{0.2cm}

\footnotesize

| $i$ | Binary | Reversed | Decimal $h_i$ |
|-----|--------|----------|---------------|
| 0 | 0 | 0 | 0.0 |
| 1 | 1 | 1 | 0.5 |
| 2 | 10 | 01 | 0.25 |
| 3 | 11 | 11 | 0.75 |
| 4 | 100 | 001 | 0.125 |
| 5 | 101 | 101 | 0.625 |

\normalsize

**Pattern:** Each point bisects the largest remaining gap

## Halton Jitter - Mathematical Formulation

**For observation $i$ in tie group:**

$$h_i = \text{VanDerCorput}(i, \text{base} = 2)$$

$$\text{displacement}_i = \epsilon \times (h_i - 0.5)$$

\vspace{0.2cm}

**Centering around 0.5** creates symmetric bidirectional displacement

\vspace{0.2cm}

**Theoretical Guarantees - Discrepancy Theory:**

Star discrepancy measures uniformity [@niederreiter1992]:

$$D_n^* = \sup_{I \subseteq [0,1]} \left| \frac{\#\{x_i \in I\}}{n} - |I| \right|$$

- Random sequences: $D_n = O(n^{-1/2})$
- Halton sequences: $D_n = O(n^{-1} \log n)$ ← near-optimal
- Optimal lower bound: $D_n = \Omega(n^{-1} \log n)$

## Halton Jitter - Applications

**Widely Used in:**

\vspace{0.2cm}

- **Quasi-Monte Carlo integration**: Better convergence than random sampling
- **Computer graphics**: Anti-aliasing, global illumination
- **Ray tracing**: Sample generation for realistic rendering
- **Numerical analysis**: Multidimensional quadrature
- **Machine learning**: Hyperparameter search spaces

\vspace{0.2cm}

**Higher-Dimensional Extensions:**

For 2D applications (e.g., scatter plots):

- $x_i = \text{VanDerCorput}(i, 2)$ (base 2 for x-axis)
- $y_i = \text{VanDerCorput}(i, 3)$ (base 3 for y-axis)

Different prime bases for each dimension maintain low-discrepancy

## Method 3: Intelligent Jitter - Novel Exploration

**Design Motivation:**

Research question: Can we apply golden ratio directly in 1D rather than through angular spacing?

\vspace{0.2cm}

**Mathematical Formulation:**

For observation $j$ in tie group of size $n_{\text{ties}}$:

$$\text{angle}_j = (j-1) \times 2\pi \times 0.618$$

$$\text{displacement}_j = \epsilon \times \cos(\text{angle}_j) \times \frac{j-1}{n_{\text{ties}}}$$

\vspace{0.2cm}

**Key Distinctions from Sunflower:**

- Angle: Golden ratio × $2\pi$ (224.4°) vs. Golden angle (137.5°)
- Scaling: **Linear** $(j/n)$ vs. Square root $(\sqrt{j/n})$
- Projection: 1D cosine modulation vs. 2D spiral $\rightarrow$ 1D

## Intelligent Jitter - Failure Analysis

**Initial Hypothesis:**

Linear scaling would create "progressive reveal":

- Early observations: Small displacement (near true value)
- Later observations: Larger displacement (fill space)
- Intuitive interpretation: "early arrivals" cluster, "late arrivals" spread

\vspace{0.2cm}

**Empirical Reality: Three Critical Problems**

**Problem 1: Excessive Displacement**

For $n = 100$:

- Observation 1: 0% of $\epsilon$
- Observation 50: 49% of $\epsilon$
- Observation 100: 99% of $\epsilon$ ← Near boundary!

## Why Intelligent Jitter Fails

**Problem 2: Artificial Stratification**

Linear scaling creates visible "layers" that don't represent data structure

- Visual appearance suggests distinct sub-groups
- These "clusters" are algorithmic artifacts
- Misrepresents uniform density as stratified distribution

\vspace{0.2cm}

**Problem 3: Perceptual Distortion**

Users interpret visual patterns as data patterns:

- Large gaps appear meaningful (but are artifacts)
- Density gradients suggest ordering (observations are exchangeable)
- Boundary concentration implies separation (all values are tied)

\vspace{0.2cm}

**Root Cause:**

Linear scaling violates uniformity and perceptual validity principles

## Value of This Negative Result

**Scientific Contributions:**

\vspace{0.2cm}

1. **Design Pattern to Avoid**
   - Lesson: Linear displacement scaling creates misleading stratification
   - Implication: Future methods should use constant or sublinear scaling

2. **Golden Ratio Not Universal**
   - Lesson: Works in specific geometric contexts, not universally
   - Implication: Biomimetic approaches require careful adaptation

3. **Importance of Scaling Function**
   - Lesson: Scaling function as critical as distribution algorithm
   - Implication: Must consider angular distribution AND radial scaling together

4. **Empirical Validation Essential**
   - Lesson: Theoretical elegance ≠ practical effectiveness
   - Implication: User studies necessary even for mathematically motivated methods

# Comparative Analysis

## Dimensional Analysis

\footnotesize

| Method | Approach | Dimension | Projection |
|--------|----------|-----------|------------|
| **Halton** | Pure 1D sequence | 1D | None |
| **Sunflower** | 2D spiral | 2D $\rightarrow$ 1D | Cosine |
| **Intelligent** | 1D with 2D-inspired modulation | Hybrid | Cosine |

\normalsize

## Scaling Behavior

\footnotesize

| Method | Scaling Function | Growth Rate | At $n=50$, $j=25$ |
|--------|-----------------|-------------|-------------------|
| **Halton** | Uniform distribution | Constant | ~$0.5\epsilon$ |
| **Sunflower** | $\sqrt{j/n}$ | Sublinear | ~$0.7\epsilon$ |
| **Intelligent** | $j/n$ | Linear | ~$0.5\epsilon$ |

\normalsize

\vspace{0.2cm}

**Distribution Quality Metrics:**

- **Minimum separation**: Halton guaranteed $O(1/n)$, highly predictable
- **Discrepancy**: Halton $O(n^{-1} \log n)$ near-optimal
- **Visual clustering**: Halton minimal, Sunflower slight central concentration, Intelligent severe stratification

## Use Case Recommendations

**Halton: Best for...**

- Precision-critical applications (scientific publications)
- Maximum uniformity requirements
- Mathematical rigor and provable guarantees
- Large datasets with efficient performance

\vspace{0.2cm}

**Sunflower: Best for...**

- General purpose use (good balance of properties)
- Aesthetic presentations
- Exploratory analysis
- User preference (often preferred in studies)

\vspace{0.2cm}

**Intelligent: Best for...**

- Methodological research (comparison baseline)
- Educational examples (teaching what NOT to do)
- **DO NOT use for production visualizations**

# Integration with ggpcp Package

## Architectural Integration

**ggpcp's Three Core Modules:**

1. **Variable Selection** (`pcp_select`): Choose and order dimensions
2. **Axis Scaling** (`pcp_scale`): Normalize or transform scales
3. **Tie Resolution** (`pcp_arrange`): Handle overlapping values $\leftarrow$ **EXTENDED**

\vspace{0.2cm}

**Our Contribution:**

Extends `pcp_arrange` to handle numerical ties alongside existing categorical tie-breaking

\vspace{0.2cm}

**Design Philosophy:**

- Maintains backward compatibility
- Follows Grammar of Graphics principles
- Integrates seamlessly with existing workflow

## Proposed ggpcp Implementation

**Function Signature:**

```r
pcp_arrange(
  data,
  method = c("from-left", "from-right", 
             "halton", "sunflower", "intelligent"),
  space = 0.05,
  epsilon = NULL,
  numeric_ties = TRUE
)
```

\vspace{0.2cm}

**New Parameters:**

- `method`: Now includes "halton", "sunflower", "intelligent"
- `epsilon`: Maximum displacement for numerical ties
  - `NULL` (default): Auto-determined as 0.05 × axis range
  - Numeric value: User-specified displacement
- `numeric_ties`: Whether to apply jittering (default: `TRUE`)

## Example Usage

```r
library(ggpcp)
library(dplyr)

# Basic usage with Sunflower (recommended default)
iris_plot <- iris %>%
  pcp_select(1:5) %>%
  pcp_scale(method = "uniminmax") %>%
  pcp_arrange(method = "sunflower") %>%
  ggplot() +
  geom_pcp()

# Halton for maximum uniformity
iris_halton <- iris %>%
  pcp_select(Sepal.Length:Species) %>%
  pcp_arrange(
    method = "halton",
    epsilon = 0.08,
    numeric_ties = TRUE
  ) %>%
  ggplot() +
  geom_pcp(aes(color = Species))
```

## Mixed Categorical and Numerical Ties

```r
# Handle both categorical and numerical ties
mixed_data %>%
  pcp_select(cat1, num1, cat2, num2) %>%
  pcp_arrange(
    method = "sunflower",  # Applied to numerical
    space = 0.05           # Applied to categorical
  )

# Method comparison
library(patchwork)

p_none <- iris %>% pcp_select(1:4) %>%
  pcp_arrange(method = "none") %>% plot_pcp()

p_halton <- iris %>% pcp_select(1:4) %>%
  pcp_arrange(method = "halton") %>% plot_pcp()

p_sunflower <- iris %>% pcp_select(1:4) %>%
  pcp_arrange(method = "sunflower") %>% plot_pcp()

(p_none | p_halton | p_sunflower) +
  plot_annotation(
    title = "Comparison of Tie-Breaking Methods"
  )
```

## Documentation Requirements

**Function Documentation:**

- Detailed explanation of each method
- Theoretical foundations and references
- When to use each method
- Parameter selection guidance
- Examples with multiple datasets

\vspace{0.2cm}

**Vignettes:**

1. "Handling Numerical Ties in ggpcp"
2. "Comparing Tie-Breaking Methods"
3. "Advanced Tie Resolution"
4. "Theory of Deterministic Jittering"

\vspace{0.2cm}

**Visual Indicators:**

Optional indicators showing:

- Which axes have tie-breaking applied
- Magnitude of epsilon used
- Number of tied observations per group

# Research Questions and Methodology

## Primary Research Question

\Large

**How can the formal structure of the Grammar of Graphics be extended to systematically incorporate and evaluate methods for resolving numerical ties in parallel coordinate plots, and what is the quantifiable impact of these methods on the accuracy and efficiency of visual data analysis?**

\normalsize

## RQ1: Theory

**How can the management of numerical ties be most effectively and coherently formalized within the layered grammar of graphics, building on the established ggpcp framework?**

\vspace{0.2cm}

**Methodology:**

- Theoretical analysis of Grammar of Graphics structure
- Literature synthesis on position adjustments
- Specification of new grammatical element
- Integration with existing ggpcp architecture
- Formal documentation of tie-breaking grammar

\vspace{0.2cm}

**Deliverables:**

- Formal specification document
- Extended grammar notation
- Theoretical paper on biomimetic transformations
- Integration guidelines for ggpcp

## RQ2: Methodology

**What are the optimal algorithmic criteria for ordering and spacing tied data points to maximize visual clarity while preserving underlying data properties?**

\vspace{0.2cm}

**Methodology:**

- Algorithm design and implementation in R
- Comparative analysis of distribution quality
- Computational performance benchmarking
- Parameter sensitivity analysis
- Edge case identification and handling

\vspace{0.2cm}

**Evaluation Metrics:**

- Minimum separation distance
- Discrepancy (uniformity measure)
- Computational complexity
- Memory efficiency
- Scalability testing

## RQ3: Perception

**How do different visualization strategies for numerical ties affect an analyst's ability to perform key visual tasks?**

\vspace{0.2cm}

**Study Design:**

- Type: Within-subjects repeated measures
- Participants: 30-50 analysts (mixed expertise)
- Methods: No jitter, Random, Halton, Sunflower, Intelligent
- Tasks:
  - Density estimation
  - Cluster identification
  - Outlier detection
  - Pattern tracing

\vspace{0.2cm}

**Dependent Variables:**

1. Accuracy (absolute error from ground truth)
2. Completion time (seconds)
3. Confidence (self-reported 1-10 scale)
4. Preference (comparative ranking)

## RQ3: Expected Hypotheses

**Statistical Analysis:**

- Repeated-measures ANOVA
- Bonferroni post-hoc tests
- Effect size calculations (partial $\eta^2$)
- Correlation analysis (accuracy vs. confidence)

\vspace{0.2cm}

**Expected Hypotheses:**

- **H1**: Halton and Sunflower > Random > No Jitter (accuracy)
- **H2**: Halton ≈ Sunflower < Random < No Jitter (time)
- **H3**: Sunflower ≈ Halton > Random > No Jitter (preference)
- **H4**: Intelligent performs poorly across all metrics

## RQ4: Practice

**Can a set of evidence-based heuristics be developed to guide practitioners in selecting the most appropriate numerical tie-breaking method for their specific data context?**

\vspace{0.2cm}

**Methodology:**

- Synthesize findings from RQ1-3
- Develop decision tree/flowchart
- Validate with case studies
- Gather practitioner feedback
- Refine through iterative testing

\vspace{0.2cm}

**Case Study Domains:**

1. Bioinformatics: Gene expression data
2. Finance: Market data with discrete prices
3. Engineering: Sensor data with limited precision
4. Social Science: Survey responses with Likert scales
5. Climate Science: Model ensemble outputs

# Implementation Roadmap

## Phase 1: Algorithm Refinement (Winter 2025)

**Tasks:**

1. Finalize all three jittering implementations
2. Develop adaptive epsilon selection
3. Optimize computational performance
4. Complete test suite with edge cases
5. Benchmark against large datasets

\vspace{0.2cm}

**Deliverables:**

- Optimized R functions
- Unit tests with 100% coverage
- Performance benchmarks
- Technical documentation

## Phase 2: ggpcp Integration (Spring 2026)

**Tasks:**

1. Extend `pcp_arrange()` function
2. Implement automatic tie detection
3. Add epsilon auto-determination
4. Create visual indicators
5. Write package vignettes
6. Prepare for CRAN submission

\vspace{0.2cm}

**Deliverables:**

- Updated ggpcp package
- Comprehensive documentation
- Three tutorial vignettes
- Package ready for CRAN

## Phase 3: User Study (Spring-Summer 2026)

**Tasks:**

1. Obtain IRB approval (early Spring)
2. Develop study materials
3. Recruit participants
4. Conduct study sessions
5. Analyze results
6. Write empirical paper

\vspace{0.2cm}

**Deliverables:**

- IRB approval documentation
- Complete dataset
- Statistical analysis
- Empirical research paper draft

## Phase 4: Case Studies & Writing (Summer 2026)

**Tasks:**

1. Apply to diverse real-world datasets
2. Gather practitioner feedback
3. Develop decision heuristics
4. Write dissertation chapters
5. Integrate all components
6. Prepare defense presentation

\vspace{0.2cm}

**Deliverables:**

- Five domain case studies
- Practitioner's guide
- Complete dissertation draft
- Defense presentation

## Phase 5: Final Review and Defense (May-July 2026)

**Tasks:**

1. Committee review of dissertation
2. Incorporate feedback
3. Final revisions
4. Defense rehearsals
5. Dissertation defense

\vspace{0.2cm}

**Deliverables:**

- Final dissertation
- Successful defense
- Submitted for graduation

# Expected Outcomes

## Theoretical Contributions

**1. Grammar of Graphics Extension**

- Formal specification of biomimetic transformations
- Integration of natural optimization principles
- New category of position adjustments

\vspace{0.2cm}

**2. Cross-Domain Algorithm Adaptation**

- Phyllotaxis $\rightarrow$ data visualization
- Quasi-random sequences $\rightarrow$ statistical graphics
- Demonstrates value of interdisciplinary approaches

\vspace{0.2cm}

**3. Negative Result Documentation**

- Intelligent jitter failure analysis
- Design patterns to avoid
- Methodological lessons for future research

## Methodological Contributions

**1. Three Novel Algorithms**

- Halton jitter for PCPs
- Sunflower jitter for PCPs
- Intelligent jitter (with failure analysis)

\vspace{0.2cm}

**2. Comparative Framework**

- Systematic evaluation criteria
- Quantitative metrics
- Perceptual assessment methods

\vspace{0.2cm}

**3. Implementation Quality**

- Production-ready R code
- Comprehensive testing
- Extensive documentation

## Practical Contributions

**1. ggpcp Package Enhancement**

- Complete tie-handling solution
- Categorical + numerical ties
- Unified grammar interface

\vspace{0.2cm}

**2. User Guidance**

- Evidence-based selection heuristics
- Interactive decision tools
- Tutorial materials

\vspace{0.2cm}

**3. Real-World Impact**

- Improved exploratory data analysis
- More accurate pattern detection
- Better density visualization

## Empirical Contributions

**1. User Study Results**

- Quantitative performance data
- Perceptual effectiveness measures
- Preference rankings

\vspace{0.2cm}

**2. Case Study Collection**

- Diverse domain applications
- Best practices examples
- Common pitfall documentation

\vspace{0.2cm}

**3. Benchmark Dataset**

- Performance comparisons
- Scalability testing
- Reference implementations

# Broader Implications

## Beyond Parallel Coordinates

**The methods generalize to other visualization contexts:**

\vspace{0.2cm}

**1. 2D Scatter Plots**

Problem: Overplotting with tied values

Solution: Full 2D Sunflower and 2D Halton

\vspace{0.2cm}

**2. Time Series Visualization**

Problem: Multiple series with identical values at time points

Solution: Vertical displacement using deterministic methods

\vspace{0.2cm}

**3. Network Visualization**

Problem: Node positioning with spatial constraints

Solution: Optimal space-filling using golden angle principles

## General Principle

\Large

**Wherever random jitter is currently used, deterministic low-discrepancy alternatives should be considered.**

\normalsize

\vspace{0.2cm}

**Benefits:**

- Reproducibility for scientific publications
- Better distribution quality
- Elimination of clustering artifacts
- Theoretical guarantees on uniformity

\vspace{0.2cm}

**Broader Impact:**

Establishes design patterns applicable across visualization domains and establishes principles for perceptually-valid position adjustments

# Timeline

## Timeline to Dissertation Defense

\footnotesize

| Phase | Timeframe | Key Milestones |
|-------|-----------|----------------|
| **Algorithm Refinement** | Winter 2025 (Months 1-2) | Algorithm optimization, adaptive epsilon |
| **ggpcp Integration** | Spring 2026 (Months 3-4) | Package update, documentation |
| **User Study** | Spring-Summer 2026 (Months 5-7) | IRB approval, data collection, analysis |
| **Case Studies & Writing** | Summer 2026 (Months 7-8) | Real-world validation, dissertation drafting |
| **Dissertation Completion** | May-June 2026 | Final revisions, committee review |
| **Defense** | **July 2026** | **Final defense and submission** |

\normalsize

# Conclusion

## Summary of Contribution

**The Problem:**

Numerical ties in parallel coordinate plots create severe visual occlusion, preventing:

- Density visualization
- Individual observation tracing
- Cluster identification
- Pattern detection

\vspace{0.2cm}

**The Solution:**

Systematic approach using three deterministic jittering methods:

- **Halton**: Quasi-random sequences with mathematical guarantees
- **Sunflower**: Biomimetic approach leveraging natural optimization
- **Intelligent**: Exploratory method demonstrating negative results

\vspace{0.2cm}

**The Impact:**

Complete framework for tie resolution in PCPs, extending Grammar of Graphics and enabling reproducible, high-quality visualizations

## Key Findings

**Evidence-Based Conclusions:**

\vspace{0.2cm}

**✓ Low-discrepancy methods superior**

- Halton and Sunflower outperform random jitter
- Uniform distribution = faithful density representation
- Mathematical guarantees translate to perceptual benefits

\vspace{0.2cm}

**✓ Determinism essential**

- Reproducibility in scientific visualization [@peng2011]
- Predictable, interpretable results
- Eliminates artifacts from stochasticity

\vspace{0.2cm}

**✗ Linear scaling problematic**

- Intelligent jitter demonstrates design failure
- Excessive displacement distorts perception
- Lesson: Scaling function as critical as distribution algorithm

## Questions for Discussion

**Open Questions for Committee:**

\vspace{0.5cm}

1. Should adaptive epsilon be user-overrideable or always automatic?

2. Should package default to Sunflower or Halton?

3. Additional task types or datasets for user study?

4. Should dissertation include 2D scatter plot extension?

5. Publication strategy - single comprehensive vs. multiple focused papers?

## Acknowledgments

**Thank you:**

- Dissertation committee for guidance and feedback
- `ggpcp` package developers
- UNL Department of Statistics
- Pilot study participants
- Open-source R community

\vspace{0.5cm}

**Contact Information:**

- Email: denise.bradford@huskers.unl.edu
- GitHub: https://github.com/drbradford12/Dissertation-Data

\vspace{1cm}

\begin{center}
\Large{Questions?}
\end{center}

# References

## References {.allowframebreaks}

\footnotesize
